Fitting metabolic models to dissolved oxygen data: the Estuarine BAyesian Single-station Estimation method

### 2023-05-08 manu-draft.qmd
added comments and recommendations to discussion

The remaining three parameters $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$), $r$ (respiration, $\text{mmol}/\text{m}^3/\text{d}$), and $b$ ($\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$) are estimated by likelihood given the observed data and prior distributions for the parameters. The JAGS software [as for BASE, @Plummer03] is used with the EBASE R package [described below, @Beck22; @RCT22] to estimate the unknown parameters using Markov Chain Monte Carlo (MCMC) simulations in a Bayesian framework. At each time step, the change in oxygen concentration between time steps is calculated from the equation using model inputs and parameter guesses based on the prior distributions, and then a finite difference approximation is used to estimate modeled oxygen concentration. Estimations proceed using a forward in time integration where the dissolved oxygen at each time step is informed by parameter estimates from the preceding time steps. The estimated dissolved oxygen concentration is also returned, which can be compared to the observed values as one measure of model performance. All parameter inputs and outputs for EBASE are shown in `r run_reference(id = "tbl-ebaseiotab")`. Areal rates for $P$, $R$, and $D$ ($\text{mmol}/\text{m}^2/\text{d}$) are returned for comparability of estimates across locations where water column depth may vary.

@@

For ease of application, an R package was developed to implement the EBASE methods described above [@Beck22]. A detailed web page (<https://fawda123.github.io/EBASE/>) describes the required data inputs for estimating the metabolic parameters using the core R functions that pass inputs to the JAGS software. A JAGS model file is included with the package that implements @eq-ebase, with options to supply a custom model file as needed. Users can also specify alternative prior distributions from those in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] for each parameter. Functions are available for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D, as areal rates with units $\text{mmol}/\text{m}^2/\text{d}$), and a time series plot of the credible intervals for the $a$, $r$, and $b$ parameters. An additional option allows the user to specify the model optimization period, where the time period can vary from a minimum of one day to the maximum of the entire length of the time series. This differs from the BASE approach where the optimization period is set at one day. Sensitivity of the EBASE method to different optimization periods is described in @sec-cmpmod. All results provided herein were generated using the EBASE R package.

@@

The above analyses demonstrated that EBASE can successfully recover known metabolic parameters from a synthetic time series and provides robust estimates on real data covering a year of observations. The success of these results was predicated on the requirements of EBASE to address several critical needs that existing methods for estimating metabolism do not address for coastal applications. These needs were addressed by 1) using a Bayesian framework that estimates the best fit of the model to the data with posterior probabilities describing the likelihood of the fit, and 2) modifying the underlying mass-balance model for dissolved with several components more appropriate for estuarine settings. First, the Bayesian framework enables hypothesis testing of how a model can be fit to data by allowing the use of prior knowledge in the model fitting process, which results in posterior probabilities describing the certainty of the parameter estimates [@Hilborn13]. The advantages of the Bayesian approach have led to widespread adoption over frequentist approaches in many scientific applications, especially in the coastal and marine community [e.g., @myers2001; @borsuk2001; @brown2017], yet these methods have not been rigorously explored for estimating metabolic rates in estuaries. Our results demonstrated that even using uninformed priors, EBASE can reproduce known metabolic parameters and rates with reasonable certainty. However, an exception was the ability of the model to estimate stable values for the $b$ parameter, although the resulting estimate of gas exchange (D) was similar to the known values. This result could be explained the dominance of the relative processes that manifest signals in the dissolved oxygen time series. That is, estuarine dissolved oxygen at Apalachicola Bay may be dominated by production and respiration, and less so from gas exchange, such that the available data provided insufficient information for EBASE to estimate $b$. This result aligns with others that have suggested gas exchange may be a relatively small component of the metabolic signal in shallow estuaries [@Murrell18]. In such cases, more constrained prior distributions for $b$ could be appropriate.

The second advantage of EBASE that makes the model more appropriate for estuarine settings included several modification to the underlying mass-balance equation for dissolved oxygen (@eq-ebase). As described in @sec-basemodif, EBASE is an estuarine adaptation of BASE (@eq-base) that was developed for stream applications and many assumptions of the underlying model of the latter were inappropriate for the former. Specifically, EBASE was modified to address three primary concerns of using BASE in an estuarine setting. First, the estimation of gross primary production as a function of PAR assumed a linear relationship such that the increase of production per unit increase of PAR was consistent for any level of the latter. The BASE model includes an exponential parameter, $p$, that defines a non-linear relationship between P and PAR (although it should be noted that BASE provides an option to set $p = 1$). @Holtgrieve10 describes a simple linear photosynthesis-irradiance curve as appropriate for systems that are never light-saturated and this assumption likely holds true for estuaries that are commonly more turbid than open-ocean or freshwater locations. Second, EBASE removed the temperature-dependency of respiration and assumed that respiration did not change with water temperature, consistent with applications of the Odum method to other shallow estuaries [@Beck15; @Murrell18]. This modification addresses recent work that suggests respiration in streams may be reduced during the night with lower temperatures as the respiration of labile autochthonous organic matter may contribute to changes in the dissolved oxygen time series [@sadro2014; @schindler2017]. There is currently no justification for this assumption in estuaries. Finally, the gas exchange in EBASE used the parameterization in @Wanninkhof14 that is primarily wind-based and developed for the open-ocean, which is clearly more appropriate than the prior formulation in BASE that also includes a temperature and depth dependency. Overall, these changes also reduce the required inputs for its use and many monitoring locations nationwide include the requisite data for applying EBASE (i.e., dissolved oxygen, water temperature, salinity, PAR, and wind speed).

Our results also confirmed that the effects of tidal advection on metabolic estimates can lead to biases in understanding gross primary production and ecosystem respiration. This result has previously been shown using the Odum open water method [@kemp1980; @caffrey2003; @Beck15] and confirmed herein, whereby results using observed time series can produce over-estimates of metabolic rates with more variation than those without tidal influence. Our assessment of the synthetic time series with noise and use of the detided dissolved oyxgen time series from the observed data at Apalachicola Bay to estimate metabolism both demonstrated this bias (Figures [-@fig-synapanoscmp] and [-@fig-apacmpfig]). As a result, detiding dissolved oxygen data is recommended prior to estimating metabolism, regardless of the method used. In addition to inaccurate and imprecise metabolic estimates, tidal influences are also most likely observed as negative (or anomalous) values in the production and respiration rates using the Odum method [@caffrey2003]. Tidal effects may be less obvious using EBASE since negative values cannot be obtained based on constraints of the mass-balance model. A potential future approach for developing EBASE could be an explicit formulation of tidal advection in the model. A synoptic tidal height time series could be used in the Bayesian framework as a separate estimate of the signal of advection on the observed dissolved oxygen. This has the advantage of both detiding and estimating metabolic rates using the same model, whereas using weighted regression prior to EBASE as used herein, requires additional analysis. However, field measurements [e.g., control volume approaches, @falter2008; @nidzieko2014] may be required to provide reasonable constraints on the prior distributions for any parameters that quantify advective influences on dissolved oxygen in a Bayesian framework.

EBASE provides several computational advantages over BASE implemented in the R package that was created for the purposes herein. Most importantly, EBASE was developed to allow different model optimization lengths, whereas the BASE method was developed to estimate results for only a single day. Allowing EBASE to incorporate more than a day of observations in the optimization can likely produce more accurate and precise estimates for the metabolic parameters. Physical and biological characteristics of an ecosystem that affect the metabolic rates are likely acting at time scales longer than a day, e.g., wind patterns that can affect gas exchange as manifested in the $b$ parameter may persist for several days. Exposing the model to additional observations may produce more stable results as the parameters for the best fit are estimated by the Bayesian routine as a function of the data that are evaluated each optimization period. However, we cannot provide a precise recommendation on the number of days to use for the optimization period as the exact length of time that ecosystem characteristics can affect metabolic rates may not be known and likely varies by location. Our example using the synthetic time series from Apalachicola Bay suggests that results were similar for the 7 and 30 day optimization period. As such, using one week of data for each optimization period is likely appropriate for this location. Using one day is not recommended given the issues stated above (i.e., unstable results), whereas an optimization period longer than a month may miss important events in the time series that can affect the metabolic estimates.

The EBASE R package can also accommodate missing observations in the input data. This allows the estimation of metabolic rates for time series that cover relatively long periods of time (e.g., more than year), when data gaps are likely to occur for several reasons (e.g., equipment malfunction, routine maintenance, etc.). The R package accommodates data gaps using linear interpolation for any required parameter used by EBASE. This allows the Bayesian routine to provide continuous estimates across the time series. The results are then post-processed such that any metabolic estimates that exceed a threshold for interpolated observations are removed from the output. The analyst can determine the appropriate length of time to use for excluding results based on interpolated values. The default setting for EBASE is 12 hours, where results for an entire optimization period are removed if any interpolated period exceeded 12 hours. Although this does not apply to the synthetic time series, some of the estimates using the 2021 observed data for Apalachicola Bay were excluded from the analysis. This feature allows EBASE to be used more readily because an analyst does not need to manually screen data that include missing values.

While the EBASE theory and software provide advantages for metabolic estimates, the approach is not without limitations. Like most models, explicit choices are needed to apply EBASE to time series data. First, the use of unconstrained prior distributions was shown to reproduce known metabolic parameters with reasonable certainty, although the $b$ parameter was still relatively unstable and unconstrained priors may not provide useful estimates in estuarine settings outside of the location evaluated herein. This may require further constraining the priors, although doing so may produce biased results in the absence of additional data to justify a constraint. Alternative metabolic models could be used for informing priors, such as using the Odum method to inform the prior for the reareation coefficient, $K$, for the BASE method. Initial application of BASE to estuarine data demonstrated that unstable and unrealistic metabolic rates were returned with uninformed priors unless the $K$ parameter was constrained. A similar approach could be applied to create more accurate estimates from EBASE. An additional choice required to use EBASE is the model optimization period, although some guidance is provided above.

A final limitation of EBASE is the relatively long processing time when applying the method to time series longer than a few days. For example, the execution time applying EBASE to a year of data on a conventional laptop computer was approximately 40 minutes, depending on the model settings. This time is compared to the Odum method applied to the same dataset, where results are obtained in a matter of seconds. Given the Bayesian framework used by EBASE, not considering the additional statistical advantages, it is not reasonable to expect comparable processing times between the methods. Regardless, computation times could be reduced with additional hardware improvements (e.g., accessing multiple processing cores) or changing the model estimation methods (e.g., reducing the number of MCMC simulations used by JAGS), although the latter may negatively affect model accuracy.

@@

-   other main conclusions (from RGN comments on draft)

>>>>

The remaining three parameters $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$), $r$ (respiration, $\text{mmol}/\text{m}^3/\text{d}$), and $b$ ($\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$) are estimated by likelihood given the observed data and prior distributions for the parameters. The JAGS software [as for BASE, @Plummer03] is used with the EBASE R package [described below, @Beck23; @RCT22] to estimate the unknown parameters using Markov Chain Monte Carlo (MCMC) simulations in a Bayesian framework. At each time step, the change in oxygen concentration between time steps is calculated from the equation using model inputs and parameter guesses based on the prior distributions, and then a finite difference approximation is used to estimate modeled oxygen concentration. Estimations proceed using a forward in time integration where the dissolved oxygen at each time step is informed by parameter estimates from the preceding time steps. The estimated dissolved oxygen concentration is also returned, which can be compared to the observed values as one measure of model performance. All parameter inputs and outputs for EBASE are shown in `r run_reference(id = "tbl-ebaseiotab")`. Areal rates for $P$, $R$, and $D$ ($\text{mmol}/\text{m}^2/\text{d}$) are returned for comparability of estimates across locations where water column depth may vary.

@@

For ease of application, an R package was developed to implement the EBASE methods described above [@Beck23]. A detailed web page (<https://fawda123.github.io/EBASE/>) describes the required data inputs for estimating the metabolic parameters using the core R functions that pass inputs to the JAGS software. A JAGS model file is included with the package that implements @eq-ebase, with options to supply a custom model file as needed. Users can also specify alternative prior distributions from those in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] for each parameter. Functions are available for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D, as areal rates with units $\text{mmol}/\text{m}^2/\text{d}$), and a time series plot of the credible intervals for the $a$, $r$, and $b$ parameters. An additional option allows the user to specify the model optimization period, where the time period can vary from a minimum of one day to the maximum of the entire length of the time series. This differs from the BASE approach where the optimization period is set at one day. Sensitivity of the EBASE method to different optimization periods is described in @sec-cmpmod. All results provided herein were generated using the EBASE R package.

@@

The above analyses demonstrated that EBASE can successfully recover known metabolic parameters from a synthetic time series and provides robust estimates on real data covering a year of observations. The success of these results was predicated on the requirements of EBASE to address several critical needs that existing methods for estimating metabolism do not address for coastal applications. These needs were addressed by 1) using a Bayesian framework that estimates the best fit of the model to the data with posterior probabilities describing the likelihood of the fit, and 2) modifying the underlying mass-balance model for dissolved oxygen with several components more appropriate for estuarine settings. First, the Bayesian framework enables hypothesis testing of how a model can be fit to data by using prior knowledge in the model fitting process, which results in posterior probabilities describing the certainty of the parameter estimates [@Hilborn13]. The advantages of the Bayesian approach have led to widespread adoption over frequentist approaches in many scientific applications, especially in the coastal and marine community [e.g., @myers2001; @borsuk2001; @brown2017], yet these methods have not been rigorously explored for estimating metabolic rates in estuaries. Our results demonstrated that even using uninformed priors, EBASE can reproduce known metabolic parameters and rates with reasonable certainty. However, an exception was the ability of the model to estimate stable values for the $b$ parameter, although the resulting estimate of gas exchange (D) was similar to the known values. This result could be explained by the dominance of the relative processes that manifest signals in the dissolved oxygen time series. That is, estuarine dissolved oxygen at Apalachicola Bay may be dominated by production and respiration, and less so from gas exchange, such that the available data provided insufficient information for EBASE to estimate $b$. This result aligns with others that have suggested gas exchange may be a relatively small component of the metabolic signal in shallow estuaries [@Murrell18]. In such cases, more constrained prior distributions for $b$ could be appropriate or EBASE could be tested in settings with larger contributions of gas exchange expressed by dissolved oxygen.

The second advantage of EBASE that makes the model more appropriate for estuarine settings included several modification to the underlying mass-balance equation for dissolved oxygen (@eq-ebase). As described in @sec-basemodif, EBASE is an estuarine adaptation of BASE (@eq-base) that was developed for stream applications and many assumptions of the underlying model of the latter were inappropriate for the former. Specifically, EBASE was modified to address three primary concerns of using BASE in an estuarine setting. First, the estimation of gross primary production as a function of PAR assumed a linear relationship such that the increase of production per unit increase of PAR was consistent for any level of the latter. The BASE model includes an exponential parameter, $p$, that defines a non-linear relationship between P and PAR (although it should be noted that BASE provides an option to set $p = 1$). @Holtgrieve10 describes a simple linear photosynthesis-irradiance curve as appropriate for systems that are never light-saturated and this assumption likely holds true for estuaries that are commonly more turbid than open-ocean or freshwater locations. Second, EBASE removed the temperature-dependency of respiration and assumed that respiration did not change with water temperature, consistent with applications of the Odum method to other shallow estuaries [@Beck15; @Murrell18]. This modification addresses recent work that suggests respiration in streams may be reduced during the night with lower temperatures as the respiration of labile autochthonous organic matter may contribute to changes in the dissolved oxygen time series [@sadro2014; @schindler2017]. There is currently no justification for this assumption in estuaries. Finally, the gas exchange in EBASE used the parameterization in @Wanninkhof14 that is primarily wind-based and developed for the open-ocean, which is clearly more appropriate than the prior formulation in BASE that includes a temperature and depth dependency for streams. Overall, these changes also reduce the required inputs and many monitoring locations nationwide include the requisite data for applying EBASE (i.e., dissolved oxygen, water temperature, salinity, PAR, and wind speed).

Our results also confirmed that the effects of tidal advection on metabolic estimates can lead to biases in understanding gross primary production and ecosystem respiration. This result has previously been shown using the Odum open water method [@kemp1980; @caffrey2003; @Beck15] and confirmed herein, whereby results using observed time series can produce over-estimates of metabolic rates with more variation than those without tidal influence. Our assessment of the synthetic time series with noise and use of the detided dissolved oyxgen time series from the observed data at Apalachicola Bay to estimate metabolism both demonstrated this bias (Figures [-@fig-synapanoscmp] and [-@fig-apacmpfig]). As a result, detiding dissolved oxygen data is recommended prior to estimating metabolism, regardless of whether EBASE or an alternative method is used. In addition to inaccurate and imprecise metabolic estimates, tidal influences are also most likely observed as negative (or anomalous) values in the production and respiration rates using the Odum method [@caffrey2003]. Tidal effects may be less obvious using EBASE since negative values cannot be obtained based on constraints of the prior distributions, although as noted above, larger and more variable estimates are returned. A potential future approach for developing EBASE could be an explicit formulation of tidal advection in the model using a synoptic tidal height time series in the Bayesian framework. This has the advantage of both detiding and estimating metabolic rates using the same model, whereas using weighted regression prior to EBASE as used herein, requires additional analysis and assumptions. However, field measurements [e.g., control volume approaches, @falter2008; @nidzieko2014] may be required to provide reasonable constraints on the prior distributions for any parameters that quantify advective influences on dissolved oxygen.

EBASE provides several computational advantages over BASE implemented in the R package that was created for the purposes herein [@Beck23]. Most importantly, EBASE was developed to allow different model optimization lengths, whereas the BASE method was developed to estimate results for a single day. Allowing EBASE to incorporate more than a day of observations in the optimization can likely produce more accurate and precise estimates for the metabolic parameters. Physical and biological characteristics of an ecosystem that affect the metabolic rates are likely acting at time scales longer than a day, e.g., wind patterns that can affect gas exchange as quantified by the $b$ parameter may persist for several days. Exposing the model to additional observations may produce more stable results as the parameters for the best fit are estimated by the Bayesian routine as a function of the data that are evaluated each optimization period. However, we cannot provide a precise recommendation on the number of days to use for the optimization period as the exact length of time that ecosystem characteristics can affect metabolic rates may not be known and likely varies by location. Our example using the synthetic time series from Apalachicola Bay suggests that results were similar for the 7 and 30 day optimization period. As such, using one week of data for each optimization period is likely appropriate for this location. Using one day is not recommended given the issues stated above (i.e., unstable results), whereas an optimization period longer than a month may miss important events in the time series that can affect the metabolic estimates.

The EBASE R package can also accommodate missing observations in the input data. This allows the estimation of metabolic rates for time series that cover relatively long periods of time (e.g., more than year), when data gaps are likely to occur for several reasons (e.g., equipment malfunction, routine maintenance, etc.). The R package accommodates data gaps using linear interpolation for any required parameter used by EBASE. This allows the Bayesian routine to provide continuous estimates across the time series. The results are then post-processed such that any metabolic estimates that exceed a threshold for interpolated observations are removed from the output. The analyst can determine the appropriate length of time to use for excluding results based on interpolated values. The default setting for EBASE is 12 hours, where results for an entire optimization period are removed if any interpolated period exceeded 12 hours within that period. Although this does not apply to the synthetic time series, some of the estimates using the 2021 observed data for Apalachicola Bay were excluded from the analysis. This feature allows EBASE to be applied more easily to other data setes because an analyst does not need to manually screen data that include missing values.

While the EBASE theory and software provide advantages for metabolic estimates, the approach is not without limitations. Like most models, explicit choices are needed to apply EBASE to time series data. First, the use of unconstrained prior distributions was shown to reproduce known metabolic parameters with reasonable certainty, although the $b$ parameter was still relatively unstable and unconstrained priors may not provide useful estimates in estuarine settings outside of the location evaluated herein. This may require further constraining the priors, although doing so may produce biased results in the absence of additional data to justify a constraint. Alternative metabolic models could be used for informing priors, such as using the Odum method to inform the prior for the reareation coefficient, $K$, for the BASE method. Initial application of BASE to estuarine data demonstrated that unstable and unrealistic metabolic rates were returned with uninformed priors unless the $K$ parameter was constrained. A similar approach could be applied to create more accurate estimates from EBASE depending on location. An additional choice for using EBASE is the model optimization period, although some guidance is provided above.

A final limitation of EBASE is the relatively long processing time when applying the method to a time series longer than a few days. For example, the execution time applying EBASE to a year of data on a conventional laptop computer was approximately 40 minutes, depending on the model settings. This time is compared to the Odum method applied to the same dataset, where results are obtained in a matter of seconds. Given the Bayesian framework used by EBASE, not considering the additional statistical advantages, it is not reasonable to expect comparable processing times between the methods. Regardless, computation times could be reduced with additional hardware improvements (e.g., accessing multiple processing cores) or changing the model estimation methods (e.g., reducing the number of MCMC simulations used by JAGS), although the latter may negatively affect model accuracy.

@@

The development of EBASE represents a new approach for estimating metabolic rates in estuarine settings that leverages existing methods by adapting theoretical components of the underlying mass-balance model using a Bayesian framework. Our results have demonstrated that unknown metabolic parameters can be estimated with reasonable certainty using uninformed prior distributions and EBASE could be a valuable approach to improve the understanding of critical ecosystem processes as the method can be readily applied to other settings. A notable limitation of EBASE was an inability to preciseily recover the $b$ parameter from a synthetic time series, although this result may be explained by a minimal contribution of gas exchange as expressed in the dissolved oxygen time series. Effects of tidal advection on dissolved oxygen measurements were also observed, demonstrating a violation of the assumption that the measured data represent a water sample with the same metabolic history. Further development of EBASE to accommodate a tidal signal, or exploring alternative methods for detiding dissolved oxygen data, represent promising avenues of future research to more accurately quantify metabolic parameters in estuarines. Regardless, EBASE represents a potentially powerful tool that removes many of the assumptions of existing methods by allowing unknown metabolic parameters to be estimated by the data and can also incorporate existing knowledge for further exploration of estuarine ecosystem properties.
